---
title: "4. Routing"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Routing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(global.par = TRUE)
```

```{r plot, echo=FALSE, results='asis'}
# plot margins
par(mar = c(1, 1, 1, 1))
# crayon needs to be explicitly activated in Rmd
options(crayon.enabled = TRUE)
# Hooks needs to be set to deal with outputs
# thanks to fansi logic
old_hooks = fansi::set_knit_hooks(
  knitr::knit_hooks,
  which = c("output", "message", "error")
)
```

Calculating shortest paths between pairs of nodes is a core task in network analysis. The `sfnetworks` package offers wrappers around the path calculation functions of `igraph`, making it easier to use them when working with spatial data and tidyverse packages. This vignette demonstrates their functionality.

In this regard it is important to remember that `sfnetworks` is a general-purpose package for spatial network analysis, not specifically optimized for a single task. If your *only* purpose is many-to-many routing in large networks, there might be other approaches that are faster and fit better to your needs. For example, the [dodgr](https://github.com/ATFutures/dodgr) package was designed for many-to-many routing on large dual-weighted graphs, with its main focus on OpenStreetMap road network data. The [cppRouting](https://github.com/vlarmet/cppRouting) package contains functions to calculate shortest paths and isochrones/isodistances on weighted graphs. When working with OpenStreetMap data, it is also possible to use the interfaces to external routing engines such as [graphhopper](https://github.com/crazycapivara/graphhopper-r), [osrm](https://github.com/rCarto/osrm) and [opentripplanner](https://github.com/ropensci/opentripplanner). The [stplanr](https://github.com/ropensci/stplanr) package for sustainable transport planning lets you use many routing engines from a single interface, through `stplanr::route()`, including routing using local R objects. Of course, all these packages can be happily used *alongside* `sfnetworks`.

```{r, message=FALSE}
library(sfnetworks)
library(sf)
library(tidygraph)
library(tidyverse)
```

## Setting edge weights

The route of the shortest path between two nodes in a network is influenced by the *weights* or *impedance* of the edges. If you don't provide these weights explicitly, shortest paths calculation functions in `sfnetworks` proceed as follows:

- If there is a column in the edges table named *weight*, the values in this column are automatically used as edge weights. These values are meant to be numeric. If they are not, `igraph` will silently try to convert them with `as.numeric()`.
- If there is no *weight* column, `st_network_paths()` internally calculates the geographic lengths of the edges and uses those as weights.

Every function that is about the calculation of shortest paths also accepts a `weights` argument which allows you to provide edge weights explicitly. In `sfnetworks`, this argument can be a numeric vector of the same length as the number of edges in the network, but also the name of a column in the edges table. The `weights` argument will always overrule the default behavior of using the *weight* column for edge weights. If you have a *weight* column but want to calculate paths *without* using weights, set `weights = NA`. In practice that means that the weight of all edges is equal to 1. Hence, the shortest path between node A and node B is the path with the fewest *number* of edges.

The `edge_length()` function can be used to calculate the length of each edge in geographic space, such that you can use those as weights in shortest paths calculations. The function is meant to be used inside a `dplyr::mutate()` verb, as follows:

```{r}
net = as_sfnetwork(roxel, directed = FALSE) %>%
  st_transform(3035) %>%
  activate("edges") %>%
  mutate(weight = edge_length())

net
```

## Calculating shortest paths

The function `st_network_paths()` is a wrapper around the igraph function `igraph::shortest_paths()`. There are three main differences:

- Besides node indices and node names, `st_network_paths()` gives the additional option to provide any (set of) geospatial point(s) as *from* and *to* location(s) of the shortest paths, either as sf or sfc object. Provided points that do not equal any node in the network will be snapped to their nearest node before calculating the paths.
- To allow smooth integration with the tidyverse, the output of `st_network_paths()` is a tibble, with one row per returned path. The column *node_paths* contains the ordered list of node indices in the path, and the column *edge_paths* contains the ordered list of edge indices in the path.
- Whenever `weights = NULL` *and* there is no column in the edges table named *weight*, `st_network_paths()` by default uses the geographic edge length as weight.

Just as `igraph::shortest_paths()`, the `st_network_paths()` function is meant for one-to-one and one-to-many routing. Hence, it is only possible to provide a single *from* location, while the *to* locations can be more than one.

Lets start with the most basic example of providing node indices as *from* and *to* locations. Remember that a node index in a sfnetwork refers to the position of the node in the nodes table of the network (i.e. its row number). There is also the possibility to use character encoded node names instead of numeric node indices. This requires the nodes table to have a column *name* with a unique name for each node. 

Since we have created a *weight* column in the previous section, these weights (i.e. the length of the edge geometries) will be used automatically in the shortest paths calculation.

```{r, fig.height=5, fig.width=5}
paths = st_network_paths(net, from = 495, to = c(458, 121))
paths
paths %>%
  slice(1) %>%
  pull(node_paths) %>%
  unlist()

paths %>%
  slice(1) %>%
  pull(edge_paths) %>%
  unlist()

plot_path = function(node_path) {
  net %>%
    activate("nodes") %>%
    slice(node_path) %>%
    plot(cex = 1.5, lwd = 1.5, add = TRUE)
}

colors = sf.colors(3, categorical = TRUE)

plot(net, col = "grey")
paths %>%
  pull(node_paths) %>%
  walk(plot_path)
net %>%
  activate("nodes") %>%
  st_as_sf() %>%
  slice(c(495, 121, 458)) %>%
  plot(col = colors, pch = 8, cex = 2, lwd = 2, add = TRUE)
```

Next we will create some geospatial points that do not intersect with any node in the network. Providing them to `st_network_paths()` will first find the nearest node to each of them, and then calculate the shortest paths accordingly.

```{r, fig.height=5, fig.width=5}
p1 = st_geometry(net, "nodes")[495] + st_sfc(st_point(c(50, -50)))
st_crs(p1) = st_crs(net)
p2 = st_geometry(net, "nodes")[458]
p3 = st_geometry(net, "nodes")[121] + st_sfc(st_point(c(-10, 100)))
st_crs(p3) = st_crs(net)

paths = st_network_paths(net, from = p1, to = c(p2, p3))

plot(net, col = "grey")
paths %>%
  pull(node_paths) %>%
  walk(plot_path)
plot(c(p1, p2, p3), col = colors, pch = 8, cex = 2, lwd = 2, add = TRUE)
```

When your network consists of multiple components that are not connected to each other, it is possible that the nearest node to a provided point is located in a tiny component and only a few other nodes can be reached from it. In such cases it might be good to first reduce the network to its largest (or *n* largest) component(s) before calculating shortest paths. The tidygraph function `tidygraph::group_components()` can help with this. It assigns an integer to each node identifying the component it is in, with `1` being the largest component in the network, `2` the second largest, and so on.

```{r, fig.height=5, fig.width=5}
# Our network consists of several unconnected components.
with_graph(net, graph_component_count())

connected_net = net %>%
  activate("nodes") %>%
  filter(group_components() == 1)

plot(net, col = colors[2])
plot(connected_net, cex = 1.1, lwd = 1.1, add = TRUE)
```

Another way to calculate shortest paths, which fits nicely in the tidygraph style of working, is by using the `to_spatial_shortest_paths` morpher function. This will subset the original network to only contain those nodes and edges that appear in a shortest path between two nodes. See the [Spatial morphers](https://luukvdmeer.github.io/sfnetworks/articles/morphers.html) vignette for details.

## Retrieving a cost matrix

The shortest paths calculation as described above is only supported for one-to-one and one-to-many routing. The alternative for many-to-many routing is the calculation of a cost matrix. Instead of providing the individual paths, it returns a matrix in which entry $i,j$ is the total cost (i.e. sum of weights) of the shortest path from node $i$ to node $j$. The igraph function for this purpose is `igraph::distances()`, which in `sfnetworks` is wrapped by `st_network_cost()`, allowing again to provide sets of geospatial points as *from* and *to* locations. Note that the calculated costs refer to the paths between the *nearest nodes* of the input points. Their units are the same as the units of weights used in the calculation, in this case meters.

```{r}
st_network_cost(net, from = c(p1, p2, p3), to = c(p1, p2, p3))
```

If we don't provide any from and to points, `st_network_cost()` will by default calculate the cost matrix for the entire network.

```{r}
# Our network has 701 nodes.
with_graph(net, graph_order())

cost_matrix = st_network_cost(net)
dim(cost_matrix)
```

## Traveling salesman problem

The traveling salesman problem aims to find the shortest tour that visits a set of locations exactly ones and then returns to the starting locations. The [`TSP`](https://cran.r-project.org/web/packages/TSP/vignettes/TSP.pdf) package has implemented a way to solving this problem through R. The input needed for the `TSP` package is a distance or costs matrix. When combining `TSP` and `sfnetworks`, we can get a solution to the traveling salesman problem in geographical space. 

First we can generate a set of random points within our network bounding box that will represent the locations we want to visit. 

```{r}
set.seed(403)
rdm = net %>%
  st_bbox() %>%
  st_as_sfc() %>%
  st_sample(10, type = "random")
```

We can then compute a cost matrix using `st_network_cost()`, which will automatically use the `weight` columns for the calculation. We can name the rows and columns with the corresponding node indices in the network to easily interact with the output of the `TSP` package functions later on. 

```{r}
net = activate(net, "nodes")
cm = st_network_cost(net, from = rdm, to = rdm)
# Assign node indices as row and column names
rdm_on_net = st_nearest_feature(rdm, net)
row.names(cm) = rdm_on_net
colnames(cm) = rdm_on_net
```

Now, we can pass our `cm` object to the `TSP::TSP()` function to convert the cost matrix into a `TSP` object. 

```{r}
library(TSP)
tsp = TSP(cm)
tsp
```

Then, we can pass this object to the tour solver `TSP::solve_TSP()`, where we will obtain an object of class `TOUR`. This object contains internally a named vector with the order of visit for each of the locations. 

```{r}
tour = solve_TSP(tsp)
tour
idx_tour = as.numeric(names(tour))
idx_tour
```

Although TSP does not recognize the units, our tour has a length of `r round(tour_length(tour), 2)` **m**, which corresponds to the CRS unit of our network. Now that we know the order at which each location will be visited, we can compute the shortest path between each location and its subsequent one, until we reach the starting point again. We can do this by using `mapply()` and the `st_network_paths()` functions.

```{r}
from_idx = idx_tour
to_idx = c(idx_tour[2:length(idx_tour)], idx_tour[1])
l = mapply(
    st_network_paths, 
    from = from_idx, 
    to = to_idx, 
    MoreArgs = list(x = net)
  )["node_paths", ] %>%
  unlist(recursive = FALSE)
```

And finally we can visualize the solution of the traveling salesman problem in geographical space:

```{r, fig.height=5, fig.width=5}
plot(net, col = "grey")
plot(rdm, pch = 20, col = colors[2], add = TRUE)
# We can reuse the function for plotting multiple node_paths from the
# Calculating short paths section:
walk(l, plot_path)
plot(
  st_as_sf(slice(net, rdm_on_net)),
  pch = 20, col = colors[3], add = TRUE
)
plot(
  st_as_sf(slice(net, idx_tour[1])),
  pch = 8, cex = 2, lwd = 2, col = colors[3], add = TRUE
)
text(
  st_coordinates(st_as_sf(slice(net, idx_tour[1]))) - c(200, 90),
  labels = "start/end\npoint"
)
```

## Creating isochrones

With respect to a given point $p$ and a given travel time $t$, an isochrone is the line for which it holds that the travel time from any point on the line to or from $p$ is equal to $t$. When using distances instead of time, it is called an isodistance.

In `sfnetworks` there are no dedicated, optimized functions for calculating isochrones or isodistances. However, we can roughly approximate them by using a combination of sf and tidygraph functions. Lets first calculate imaginary travel times for each edge, using randomly generated average walking speeds for each type of edge.

```{r, warning=FALSE}
# How many edge types are there?
types = net %>%
  activate("edges") %>%
  pull(type) %>%
  unique()

types

# Randomly define a walking speed in m/s for each type,
# with values between 3 and 7 km/hr.
set.seed(1)
speeds = runif(length(types), 3 * 1000 / 60 / 60, 7 * 1000 / 60 / 60)

# Assign a speed to each edge based on its type.
# Calculate travel time for each edge based on that.
net = net %>%
  activate("edges") %>%
  group_by(type) %>%
  mutate(speed = units::set_units(speeds[cur_group_id()], "m/s")) %>%
  mutate(time = weight / speed) %>%
  ungroup()

net
```

Now, we can calculate the total travel time for each shortest path between the (nearest node of the) origin point and all other nodes in the network, using the node measure function `tidygraph::node_distance_from()` with the values in the *time* column as weights. Then, we filter the nodes reachable within a given travel time from the origin. By drawing a convex hull around these selected nodes we roughly approximate the isochrone. If we wanted isochrones for travel times *towards* the central point, we could have used the node measure function `tidygraph::node_distance_to()` instead.

```{r, fig.height=5, fig.width=5}
net = activate(net, "nodes")

p = net %>%
  st_geometry() %>%
  st_union() %>%
  st_centroid()

iso = net %>%
  filter(node_distance_from(st_nearest_feature(p, net), weights = time) <= 600)

iso_poly = iso %>%
  st_geometry() %>%
  st_union() %>%
  st_convex_hull()

plot(net, col = "grey")
plot(iso_poly, col = NA, border = "black", lwd = 3, add = TRUE)
plot(iso, col = colors[2], add = TRUE)
plot(p, col = colors[1], pch = 8, cex = 2, lwd = 2, add = TRUE)
```

## Custom routing profiles

In many cases the shortest path based geographical distances or travel time is not necessarily the optimal path. The most appropriate route may depend on many factors, for example staying away from large and potentially unpleasant highways for motor traffic. All networks have different characteristics. In OpenStreetMap networks the 'highway' type is often a good (albeit crude) approximation of the network. The `roxel` demo dataset is derived from OpenStreetMap and represents a largely residential network:

```{r}
table(roxel$type)
```

Building on the shortest paths calculated in a previous section we can try an alternative routing profile. Let's take a look at these paths to see what type of ways they travel on:

```{r, fig.height=5, fig.width=5}
paths_sf = net %>%
  activate("edges") %>%
  slice(unlist(paths$edge_paths)) %>%
  st_as_sf()

table(paths_sf$type)

plot(paths_sf["type"], lwd = 4, key.pos = 4, key.width = lcm(4))
```

As with the overall network, the paths we calculated are dominated by residential streets.
For the purposes of illustration, lets imagine we're routing for a vehicle that we want to keep away from residential roads, and which has a much lower cost per unit distance on secondary roads:

```{r}
weighting_profile = c(
  cycleway = Inf,
  footway = Inf,
  path = Inf,
  pedestrian = Inf,
  residential = 3,
  secondary = 1,
  service = 1,
  track = 10,
  unclassified = 1
)

weighted_net = net %>%
  activate("edges") %>%
  mutate(multiplier = weighting_profile[type]) %>%
  mutate(weight = edge_length() * multiplier)
```

We can now recalculate the routes. The result show routes that avoid residential networks.

```{r, fig.show='hold', out.width = '50%'}
weighted_paths = st_network_paths(weighted_net, from = 495, to = c(458, 121))

weighted_paths_sf = weighted_net %>%
  activate("edges") %>%
  slice(unlist(weighted_paths$edge_paths)) %>%
  st_as_sf()

table(weighted_paths_sf$type)

plot(st_as_sf(net, "edges")["type"], lwd = 4,
     key.pos = NULL, reset = FALSE, main = "Distance weights")
plot(st_geometry(paths_sf), add = TRUE)
plot(st_as_sf(net, "edges")["type"], lwd = 4,
     key.pos = NULL, reset = FALSE, main = "Custom weights")
plot(st_geometry(weighted_paths_sf), add = TRUE)
```

*NOTE: Developing more sophisticated routing profiles is beyond the scope of this package.
If you need complex mode-specific routing profiles, we recommend looking at routing profiles associated with open source routing engines, such as [bike.lua](https://github.com/fossgis-routing-server/cbf-routing-profiles/blob/master/bike.lua) in the OSRM project. Another direction of travel could be to extend on the approach illustrated here, but this work could be well-suited to a separate package that builds on `sfnetworks` (remembering that sophisticated routing profiles account for nodes and edges). If you'd like to work on such a project to improve mode-specific routing in R by building on this package, please let us know in the [discussion room](https://github.com/luukvdmeer/sfnetworks/discussions)!*
