---
title: "Cleaning spatial networks"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{3. Cleaning spatial networks}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto:html}
format:
  html:
    toc: true
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
  opts_knit:
    global.par: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
current_geos = numeric_version(sf::sf_extSoftVersion()["GEOS"])
required_geos = numeric_version("3.7.0")
geos37 = current_geos >= required_geos
```

```{r}
#| label: plot
#| echo: false
#| results: asis
# plot margins
oldpar = par(no.readonly = TRUE)
par(mar = c(1, 1, 1, 1))
# crayon needs to be explicitly activated in Rmd
oldoptions = options()
options(crayon.enabled = TRUE)
# Hooks needs to be set to deal with outputs
# thanks to fansi logic
old_hooks = fansi::set_knit_hooks(
  knitr::knit_hooks,
  which = c("output", "message", "error")
)
```

Unfortunately real-world datasets are not always as friendly as those used in tutorials. Cleaning the network data therefore forms a crucial part of a spatial network analysis workflow. This vignette presents some examples that may be of use when going through this phase.

```{r}
#| message: false
library(sfnetworks)
library(sf)
library(tidygraph)
library(ggraph)
library(dplyr)
```

## The basics

Functions that modify the topology of the network are implemented in `{sfnetworks}` as **spatial morphers**. Network cleaning functions belong to this family. To learn more about morphing in general, see the vignette [Spatial morphers](). For now, it is sufficient to know that you can use any spatial morpher function inside the `tidygraph::convert()` verb to convert your network into a new structure.

Before presenting the cleaning functions that are currently implemented, lets create a network to be cleaned.

```{r}
p01 = st_point(c(0, 1))
p02 = st_point(c(1, 1))
p03 = st_point(c(2, 1))
p04 = st_point(c(3, 1))
p05 = st_point(c(4, 1))
p06 = st_point(c(3, 2))
p07 = st_point(c(3, 0))
p08 = st_point(c(4, 3))
p09 = st_point(c(4, 2))
p10 = st_point(c(4, 0))
p11 = st_point(c(5, 2))
p12 = st_point(c(5, 0))
p13 = st_point(c(5, -1))
p14 = st_point(c(5.8, 1))
p15 = st_point(c(6, 1.2))
p16 = st_point(c(6.2, 1))
p17 = st_point(c(6, 0.8))
p18 = st_point(c(6, 2))
p19 = st_point(c(6, -1))
p20 = st_point(c(7, 1))

l01 = st_sfc(st_linestring(c(p01, p02, p03)))
l02 = st_sfc(st_linestring(c(p03, p04, p05)))
l03 = st_sfc(st_linestring(c(p06, p04, p07)))
l04 = st_sfc(st_linestring(c(p08, p11, p09)))
l05 = st_sfc(st_linestring(c(p09, p05, p10)))
l06 = st_sfc(st_linestring(c(p08, p09)))
l07 = st_sfc(st_linestring(c(p10, p12, p13, p10)))
l08 = st_sfc(st_linestring(c(p05, p14)))
l09 = st_sfc(st_linestring(c(p15, p14)))
l10 = st_sfc(st_linestring(c(p16, p15)))
l11 = st_sfc(st_linestring(c(p14, p17)))
l12 = st_sfc(st_linestring(c(p17, p16)))
l13 = st_sfc(st_linestring(c(p15, p18)))
l14 = st_sfc(st_linestring(c(p17, p19)))
l15 = st_sfc(st_linestring(c(p16, p20)))

lines = c(
  l01, l02, l03, l04, l05,
  l06, l07, l08, l09, l10,
  l11, l12, l13, l14, l15
)

edges = st_sf(id = seq_along(lines), geometry = lines)

net = as_sfnetwork(edges)
```

```{r}
make_ggraph = function(x) {
  ggraph(x, "sf") +
  geom_edge_sf(aes(color = as.factor(id)), linewidth = 2, show.legend = FALSE) +
  geom_node_sf(size = 4) +
  theme_void()
}

make_ggraph(net)
```

## Simplify the network

A network may contain sets of edges that connect the same pair of nodes. Such edges can be called *multiple edges*. Also, it may contain an edge that starts and ends at the same node. Such an edge can be called a *loop edge*.

In graph theory, a *simple graph* is defined as a graph that does *not* contain multiple edges nor loop edges. To obtain a simple version of our network, we can remove multiple edges and loop edges by using the morpher `to_spatial_simple()`.

```{r}
simple = net |>
  convert(to_spatial_simple)
```

```{r}
#| layout-ncol: 2
make_ggraph(net)
make_ggraph(simple)
```

By re-arranging the edges table before applying the morpher you can influence which edges are kept whenever sets of multiple edges are detected. For example, you might want to always keep the edge with the shortest distance in the set.

```{r}
simple = net |>
  activate(edges) |>
  arrange(edge_length()) |>
  convert(to_spatial_simple)
```

```{r}
#| layout-ncol: 2
make_ggraph(net)
make_ggraph(simple)
```

## Subdivide edges

When constructing a sfnetwork from a set of sf linestrings, the endpoints of those linestrings become nodes in the network. If endpoints are shared between multiple lines, they become a single node, and the edges are connected. However, a linestring geometry can also contain interior points that define the shape of the line, but are not its endpoints. It can happen that such an interior point in one edge is equal to either an interior point or endpoint of another edge. In the network structure, however, these two edges are not connected, because they don't share endpoints. If this is unwanted, we need to split these two edges at their shared point and connect them accordingly.

In graph theory terms the process of splitting and edge is called *subdivision*: the subdivision of an edge $o = \{a, b\}$ (i.e. an edge from node $a$ to node $b$) is the addition of a new node $c$ and the replacement of $o$ by two new edges $p = \{a, c\}$ and $q = \{c, b\}$.

The function `to_spatial_subdivision()` subdivides edges at interior points whenever these interior points are equal to one or more interior points or endpoints of other edges, and recalculates network connectivity afterwards.

To illustrate the workflow, lets consider a situation where an interior point $p_{x}$ in edge $x$ is shared with point $p_{y}$ in edge $y$. That gives two possible situations:

- $p_{y}$ is an interior point of $y$. Since $p_{x}$ and $p_{y}$ are both interior points, neither of them is already a node in the network. Then:
  - $x$ is subdivided at $p_{x}$ into two new edges $x_{1}$ and $x_{2}$ and new node $p_{x}$.
  - $y$ is subdivided at $p_{y}$ into two new edges $y_{1}$ and $y_{2}$ and new node $p_{y}$.
  - The new nodes $p_{x}$ and $p_{y}$ are merged into a single node $p$ with the edge set $\{x_{1}, x_{2}, y_{1}, y_{2}\}$ as incidents.
- $p_{y}$ is and endpoint of $y$. Since $p_{y}$ is an edge endpoint, it is already a node in the network. Then:
  - $x$ is subdivided at $p_{x}$ into two new edges $x_{1}$ and $x_{2}$ and new node $p_{x}$.
  - The new node $p_{x}$ is merged with node $p_{y}$ into a single node $p$ with the edge set $\{y, x_{1}, x_{2}\}$ as incidents.

Note that an edge is not subdivided when it crosses another edge at a location that is not an interior point or endpoint in the linestring geometry of any of the two edges.

For our example network, this means:

```{r}
#| warning: false
subdivision = simple |>
  convert(to_spatial_subdivision)
```

```{r}
#| layout-ncol: 2
make_ggraph(simple)
make_ggraph(subdivision)
```

### Smooth pseudo nodes

A network may contain nodes that have only one incoming and one outgoing edge. For tasks like calculating shortest paths, such nodes are redundant, because they don't represent a point where different directions can possibly be taken. Sometimes, these type of nodes are referred to as *pseudo nodes*. Note that their equivalent in undirected networks is any node with only two incident edges, since incoming and outgoing does not have a meaning there. To reduce complexity of subsequent operations, we might want to get rid of these pseudo nodes.

In graph theory terms this process is the opposite of subdivision and also called *smoothing*: smoothing a node $b$ with incident edges $o = \{a, b\}$ and $p = \{b, c\}$ removes $b$, $o$ and $p$ and creates the new edge $q = \{a, c\}$.

The function `to_spatial_smooth()` iteratively smooths pseudo nodes, and after each removal concatenates the linestring geometries of the two affected edges together into a new, single linestring geometry.

```{r}
smooth = subdivision |>
  convert(to_spatial_smooth)
```

```{r}
#| layout-ncol: 2
make_ggraph(subdivision)
make_ggraph(smooth)
```

Pseudo nodes may have a function by representing a location where a certain attribute changes. For example, from that point on the street has a different surface type. The `require_equal` argument allows you to specify one or more edge attributes that will be checked for equality before removing a pseudo node. If those attributes are equal, the node will be removed, but if not, it will remain. The argument value is evaluated using tidy selection, meaning you can specify column names unquoted, but also use [selection helpers](https://dplyr.tidyverse.org/reference/select.html).

```{r}
net_with_attrs = subdivision |>
  activate(edges) |>
  mutate(
    foo = sample(letters, n(), replace = TRUE), 
    bar = sample(c(1:10), n(), replace = TRUE)
  )

# Only smooth pseudo nodes if incident edges have the same value for "foo".
smooth_b = net_with_attrs |>
  convert(to_spatial_smooth, require_equal = foo)

# Only smooth pseudo nodes if incident edges have the same value for all attributes.
smooth_c = net_with_attrs |>
  convert(to_spatial_smooth, require_equal = everything())
```

The `to_spatial_smooth()` morpher also allows you to specify if and how you want to combine attributes of concatenated edges. This is done through the `attribute_summary` argument and explained in more detail [here]().

## Simplify intersections

Especially in road networks you may find that intersections between edges are not modeled by a single node. Instead, each leg of the intersection has a dedicated edge. To simplify the topology of your network, you might want to reduce such complex intersection structures into a single node. Hence, we want to reduce a group of nodes into a single node, while maintaining the connectivity of the network.

In graph theory terms this process is called *contraction*: the contraction of a set of nodes $P = \{p_{1}, p_{2}, ..., p_{n}\}$ is the replacement of $S$ and all its incident edges by a single node $p^{*}$ and a set of edges that connect $p^{*}$ to all nodes that were adjacent to any node $p_{i} \in P$.

The morpher `to_spatial_contracted()` contracts groups of nodes based on a given grouping variable. The geometry of each contracted node is (by default) the centroid of the original group members' geometries. Moreover, the geometries of the edges that start or end at a contracted node are updated such that their boundaries match the new node geometries.

Grouping variables are internally forwarded to `dplyr::group_by()`. That means you can group the nodes based on any (combination of) attribute(s). However, in this case, we want to group the nodes spatially, such that nodes that are very close to each other in space will form a group and get contracted. Spatial grouping of nodes can be done with the `group_spatial_dbscan()` function, which exposes the DBSCAN clustering algorithm through the `{dbscan}` package. The main parameter $\epsilon$ defines the radius of the neighborhood of each node in meters. By default, this is based on network distance rather than euclidean distance.

```{r}
contraction = smooth |>
  activate(nodes) |>
  convert(to_spatial_contracted, group_spatial_dbscan(0.5))
```

```{r}
#| layout-ncol: 2
make_ggraph(smooth)
make_ggraph(contraction)
```

The `to_spatial_contracted()` morpher also allows you to specify if and how you want to combine attributes of contracted nodes. This is done through the `attribute_summary` argument and explained in more detail [here]().

## Overview

In a single pipeline, we can clean our dirty network as follows.

```{r}
#| warning: false
clean = net |>
  activate(edges) |>
  arrange(edge_length()) |>
  convert(to_spatial_simple) |>
  convert(to_spatial_subdivision) |>
  convert(to_spatial_smooth) |>
  activate(nodes) |>
  convert(to_spatial_contracted, group_spatial_dbscan(0.5))
```

```{r}
#| layout-ncol: 2
make_ggraph(net)
make_ggraph(clean)
```

## Non-tidy workflow

All cleaning morphers have their internal worker exported as well. They can be used if you want to work outside the "tidy workflow". The same cleaning operation can be done as follows:

```{r}
#| warning: false
simple = simplify_network(net)
subdivision = subdivide_edges(simple)
smooth = smooth_pseudo_nodes(subdivision)
groups = with_graph(smooth, group_spatial_dbscan(0.5))
clean = contract_nodes(smooth, groups)
```

```{r}
#| include: false
par(oldpar)
options(oldoptions)
```
